---
title: "Real interest rates"
author: ""
format: html
---

## Download U.S. nominal interest rate

For the U.S. nominal interest rate, we use the Federal Funds Effective Rate:
<https://fred.stlouisfed.org/series/FEDFUNDS>

```{r}
library(tidyverse)
library(fredr)
fredr_set_key("xxxxx") # set key to the FRED API
# you first need to apply for a key at https://fred.stlouisfed.org/docs/api/api_key.html

nom_us = fredr(
  series_id = "FEDFUNDS",
  observation_start = as.Date("2010-01-01"),
  frequency = "m"
)
```

## Download U.S. inflation rate

For the U.S. inflation rate, we use the CPI for urban consumers:
<https://fred.stlouisfed.org/series/CPIAUCSL>

```{r}
cpi_us = fredr(
  series_id = "CPIAUCSL",
  observation_start = as.Date("2009-01-01"),
  frequency = "m"
)

# calculate 12-month percentage change in CPI
cpi_us$inf_us = cpi_us$value / lag(cpi_us$value, n=12) - 1
```

## Combine nominal interest rate and inflation rate in single dataframe:

```{r}
# drop unnecessary variables
nom_us = subset(nom_us, select = c(date, value))
cpi_us = subset(cpi_us, select = c(date, inf_us))

# rename variables
names(nom_us)[names(nom_us) == 'value'] = 'nom_us'

nom_us$nom_us = nom_us$nom_us / 100

# add the variable cpi_us to the data frame nom_us using an inner join
d = inner_join(nom_us, cpi_us, by = "date")

# calculate real interest rate using Fisher equation:
d$real_us = (1+d$nom_us) / (1+d$inf_us) - 1
```

## Download Hungarian data:

For interest rates, use "Interest Rates: Immediate Rates (< 24 Hours): Central Bank Rates: Total for Hungary" from FRED: <https://fred.stlouisfed.org/series/IRSTCB01HUM156N>

```{r}
nom_hu = fredr(
  series_id = "IRSTCB01HUM156N",
  observation_start = as.Date("2010-01-01"),
  frequency = "m"
)
```

For CPI, use "Consumer Price Indices (CPIs, HICPs), COICOP 1999: Consumer Price Index: Total for Hungary" from FRED: <https://fred.stlouisfed.org/series/CPALTT01HUM659N>

According to FRED, this is "Growth rate same period previous year, Not Seasonally Adjusted".

```{r}
cpi_hu = fredr(
  series_id = "CPALTT01HUM659N",
  observation_start = as.Date("2010-01-01"),
  frequency = "m"
)
```

## Combine nominal interest rate and inflation rate in single dataframe:

```{r}
# drop unnecessary variables
nom_hu = subset(nom_hu, select = c(date, value))
cpi_hu = subset(cpi_hu, select = c(date, value))

# rename variables
names(nom_hu)[names(nom_hu) == 'value'] = 'nom_hu'
names(cpi_hu)[names(cpi_hu) == 'value'] = 'inf_hu'

nom_hu$nom_hu = nom_hu$nom_hu / 100
cpi_hu$inf_hu = cpi_hu$inf_hu / 100

# add the variable cpi_us to the data frame nom_us using an inner join
d2 = inner_join(nom_hu, cpi_hu, by = "date")

# calculate real interest rate using Fisher equation:
d2$real_hu = (1+d2$nom_hu) / (1+d2$inf_hu) - 1

# combine d and d2 using inner join
d = inner_join(d, d2, by = "date")
```

## Plot results
```{r}
# Create a line plot using ggplot2 that shows both d$real_us and d$real_hu
ggplot(d, aes(x=date)) +
  geom_line(aes(y=real_us, color="US")) +
  geom_line(aes(y=real_hu, color="HU")) +
  labs(title="Real interest rates",
       x="Date",
       y="Real interest rate",
       color="Country") +
  theme_minimal() + 
  scale_x_date(date_breaks = "2 year", date_labels = "%Y")
```



